\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-06 chapters/contribution.tex
cover page

\begin{document}

% Front matter ================================================================
\frontmatter

\include{chapters/preface.tex}
\chapter{Abstract}
\include{chapters/abstract.tex}
\chapter{Samenvatting}
\tableofcontents
\listoffigures
\listoftables
\printacronyms[name=List of acronyms]

% Main matter =================================================================
\mainmatter%

\part{Introduction \& Literature}
\includechapter{Brain-computer interfaces}{chapters/bci.tex}
\includechapter{Gaze-independent visual BCIs}{chapters/gaze_independence.tex}
\todo{The role of jitter in gaze-independent ERP decoding}
%\includechapter{Problem statement \& rationale}{chapters/problem_statement.tex}

\part{Algorithms \& Decoders}
\includechapter{Kronecker-structured linear decoding}{chapters/stbf_struct.tex}
\includechapter{Block-term tensor discriminant analysis}{chapters/bttda.tex}
\todo{discuss Tensorizations coping with latency jitter to make link}
\includechapter{Latency estimation methods}{chapters/wcble.tex}

\part{Experiments \& Applications}
\includechapter{Compensating jitter for gaze-independce}{chapters/covert_align.tex}
\todo{Integrating gaze-tracking data}
\includechapter{Case studies}{chapters/patients.tex}
\includechapter{Conclusion \& recommendations}{chapters/conclusion.tex}

\backmatter%
\chapter{Funding \& acknowledgments}
\todo{my grant}\todo{marc's projects}
\chapter{Personal contribution}
\todo{Juliette, Alixe, Kristl Claeys, Edward Schietekatte, Marc, François,
Hakim, Marie,Etienne, Philip Van Damme  }
\chapter{Conflict of interest}
The authors declare no conflict of interest.
\include{chapters/cv.tex}

\printbibliography
\end{document}

>>>>

\chapter*{Personal contribution}
\todo{which works conceptualized}
\todo{Juliette, Alixe, Kristl Claeys, Edward Schietekatte, Marc, François,
Hakim, Marie,Etienne, Philip Van Damme  }

### 2024-10-06 chapters/gaze_independence.tex
cover page

\epigraph{%
  ``\elide\ This begs the question: if a patient had reliable control
  over their eye gaze, why would one aim to facilitate communication with that
  patient via a BCI? It would be far simpler and more efficient to use an
  eye-tracking system.
  And even if a BCI were the preferred or only available option for some reason,
  designs requiring eye gaze control could not be used
  by patients with severe motor impairment, as these patients
  cannot direct their gaze.
  Because these are the patients who have the greatest need for BCIs, it is
  important that accurate gaze-independent BCIs exist.''
}{%
  \textcite{Egan2017}
}

@@

\Ac{sspi} is often caused by damage to the central or peripheral nervous
system, either through congenital diseases (\ac{cp}, \ac{fa}, \ldots),
neurodegenerative (\ac{als}, \ac{ms}, \ldots) or acquired (stroke and
\ac{tbi}).
Many individuals in these groups unfortunately also suffer to
some extent from eye motor impairment, requiring \ac{bci}s adapted to their condition.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor impairment, which can range from minor (nystagmus
\footnote{%
Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).
}, other eye tremors
\footnote{%
These can include square-wave jerks, saccadic intrusions, microtremors, or
microsaccades while resting or fixating the gaze.
}, gaze fixation fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia
\footnote{%
Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.
}, involuntary movements, impaired pursuit, \ldots), and even complete
ophtalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only has an effect on vision and coordination but also
on their ability to operate a visual \ac{bci}~\cite{FriedOken2020}.

@@

@@

We believe it is in general not opportune to delve too deep in the underlying
ophthalmological and neurological pathophysiology for each of the specific
conditions.
In clinical reality, every patient different symptoms, which lead to their own
unique set of preserved capabilities.
From a solution-oriented \ac{bci} engineering point
of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing patients we will refer to these groups as individuals
with \ac{sspi} and, if eye motor skills are affected,  \ac{sspgi}, in line with
the terminology used by~\textcite{FrieDOken2020} (see
Figure~\ref{fig:gaze_independence/venn}).

@@

\section{State-of-the-art}
\label{sec:gaze-independence}

@@

The general goal of this PhD is to tackle the gaze-dependence in problem in the
visual oddball \ac{bci}.
Section~\ref{sec:gaze-independence} showed that visual \ac{bci} paradigms form
a good candidate for a performant gaze-independent \ac{bci}.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
visuospatial attention can be decoded separately from the position of gaze,
which do not necessarily need to coincide.

>>>>

\chapter{Gaze-independent visual \acsp{bci}}
\label{sec:gaze-independent}
\epigraph{%
  ``\elide\ This begs the question: if a patient had reliable control
  over their eye gaze, why would one aim to facilitate communication with that
  patient via a BCI? It would be far simpler and more efficient to use an
  eye-tracking system.
  And even if a BCI were the preferred or only available option for some reason,
  designs requiring eye gaze control could not be used
  by patients with severe motor impairment, as these patients
  cannot direct their gaze.
  Because these are the patients who have the greatest need for BCIs, it is
  important that accurate gaze-independent BCIs exist.''
}{%
  \textcite{Egan2017}
}

@@

\subsection{Incidence of eye motor impairment}
\Ac{sspi} is often caused by damage to the central or peripheral nervous
system, either through congenital diseases (\ac{cp}, \ac{fa}, \ldots),
neurodegenerative (\ac{als}, \ac{ms}, \ldots) or acquired (stroke and
\ac{tbi}).
Many individuals in these groups unfortunately also suffer to
some extent from eye motor impairment, requiring \ac{bci}s adapted to their condition.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor impairment, which can range from minor (nystagmus
\footnote{%
Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).
}, other eye tremors
\footnote{%
These can include square-wave jerks, saccadic intrusions, microtremors, or
microsaccades while resting or fixating the gaze.
}, gaze fixation fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia
\footnote{%
Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.
}, involuntary movements, impaired pursuit, \ldots), and even complete
ophtalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only has an effect on vision and coordination but also
on their ability to operate a visual \ac{bci}~\cite{FriedOken2020}.

@@

\subsection{Gaze impairment and \acl{lis}}

@@

We believe it is in general not opportune to delve too deep in the underlying
ophthalmological and neurological mechanisms for each of the specific
conditions, or even the exact symptoms.
In clinical reality, every patient different symptoms, which lead to their own
unique set of preserved capabilities and visual skills.
From a solution-oriented \ac{bci} engineering point
of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing patients we will refer to those that might benefit from
\ac{bci} assistive communication technology as with \ac{sspi}, in line with
the terminology used by~\textcite{FrieDOken2020}.
If they have eye motor impairment that affects their use of visual \ac{bci} or
eye tracking solutions, we use the term \ac{sspgi} (see Figure~\ref{fig:gaze_independence/venn}).

@@

\section{State-of-the-art}
\label{sec:gaze-independence/sota}

@@

The general goal of this PhD is to tackle the gaze-dependence in problem in the
visual oddball \ac{bci}.
Section~\ref{sec:gaze-independence/sota} showed that visual \ac{bci} paradigms form
a good candidate for a performant gaze-independent \ac{bci}.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
visuospatial attention can be decoded separately from the position of gaze,
which do not necessarily need to coincide.

### 2024-10-06 chapters/patients.tex
Rewrite sota
Update bttda formulation
cover page

\section{Introduction}
\todo{rename patient to bci user and participant}
\todomvh{separate terminology for sspgi}
\todo{consistently: gaze impairment, eye motor impairment or oculomotor impairment}

@@

Self-reported eye motor and visual abnormalities were recorded according to the
relevant visual \ac{bci} skills presented by~\textcite{FriedOken2020}.
These include visual acuity, visual fixation, eyelid function, ocular motility
, binocuar vision and field of vision.
Additionally, participants were asked about eye tremors (nystagmus or other) and
other involuntary eye movements.

@@

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
}{%
  Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
  conditions.
  Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
  near the middle of the screen, and PC4 near the bottom.
}{%
  fig:patients/gaze
}
\todo{explaine crosses}

>>>>

\chapter{Case studies}
\label{sec:patients}
\section{Introduction}
\todo{rename patient to bci user and participant}
\todomvh{separate terminology for sspgi}
\todo{consistently: gaze impairment, eye motor impairment or oculomotor impairment}

@@

Self-reported eye motor and visual abnormalities were recorded according to the
relevant visual \ac{bci} skills presented by~\textcite{FriedOken2020}.
These include visual acuity, visual fixation, eyelid function, ocular motility,
binocular vision and field of vision.
Additionally, participants were asked about eye tremors (nystagmus or other) and
other involuntary eye movements.

@@

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
}{%
  Distribution of the recorded gaze position.
}{%
  Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
  conditions.
  Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
  near the middle of the screen, and PC4 near the bottom.
}{%
  fig:patients/gaze
}
\todo{explaine crosses}
